---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![R-CMD-check](https://github.com/TimTaylor/surveyTools/workflows/R-CMD-check/badge.svg)](https://github.com/TimTaylor/surveyTools/actions){.pkgdown-devel}
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Overview

<span style="color:red"> ***This package is currently under active development. The code has yet been reviewed and is likely to change over the following months. We recommend it is not used for analysis for the time being.***  </span>

---

*surveyTools* aims to provide a coherent interface for the analysis of
[EQ-5D](https://euroqol.org/eq-5d-instruments) surveys (EQ-5D-3L, EQ-5D-5L and
EQ-5D-Y formats).

The package provides a range of features including:

-   Constructors (and validators) for EQ5D data frame subclasses (EQ5D3L, EQ5D5L
    and EQ5D): `new_eq5d3l()` (`validate_eq5d3l()`), `new_eq5d5l()`
    (`validate_eq5d5l()`) and `new_eq5dy()` (`validate_eq5dy()`).

-   User-friendly coercion functions `as_eq5d3l()`, `as_eq5d5l()` and
    `as_eq5dy()`.

-   The calculation of utility values based on a range of different value sets.
    This functionality is provided via the `calculate_utility()`,
    `add_utility()` and `available_valuesets()` functions which are wrappers
    around the [eq5d](https://cran.r-project.org/package=eq5d) package.

-   The calculation of different Quality of Life Years (QALY) metrics including
    unadjusted 'raw' values, and the disutility from both perfect health and,
    optionally, a specified baseline. See `calculate_qalys()`.

-   The calculation of the Paretian Classification of Health Change (PCHC) in an
    individual's health state between two surveys via `calculate_pchc()` (again
    wrapping the [eq5d](https://cran.r-project.org/package=eq5d) package).

-   Easy calculation of responses with a health limitation (i.e. a non-one
    response in one of the dimensions) via `calculate_limitation()`.

See `vignette("surveyTools")` for a gentle introduction to the main
functionality.

# Installation

The development version, which this documentation refers to, can be installed
via:


::: {.pkgdown-devel}
```{r, eval=FALSE}
    install.packages("surveyTools", repos = "https://timtaylor.r-universe.dev")
```
:::

# Getting help online

Bug reports and feature requests should be posted on *github* using the [*issue*
system](https://github.com/TimTaylor/surveyTools/issues).
