---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/TimTaylor/qalytools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/TimTaylor/qalytools/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<br>
<span style="color:red"> ***This package is currently under active development and the code subject to change.***  </span>

# qalytools

*qalytools* provides a simple and intuitive user interface for the
analysis of [EQ-5D](https://euroqol.org/eq-5d-instruments) surveys. It builds
upon the [eq5d package](https://cran.r-project.org/package=eq5d) to facilitate
the calculation of QALY metrics, and other related values, across multiple
surveys.

## Installation

The development version of the package can be installed via:
```{r, eval=FALSE}
    repos <- c("https://timtaylor.r-universe.dev", getOption("repos"))
    install.packages("qalytools", repos = repos)
```

## Guidance

Once installed the best way to familiarise yourself with the package is to view
the included documentation:

- `vignette("qalytools")` provides an introduction to the main functionality
  of the package.
- `vignette("example_analysis")` illustrates how to use the package as part of a
  larger analysis.
