---
title: "EQ-5D Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EQ-5D Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

<br>
<span style="color:red"> ***vignette under active development and subject to change.***  </span>

```{r, include = FALSE}
options(dplyr.summarise.inform = FALSE)

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    fig.width = 8,
    fig.height = 5
)

ggplot2::theme_set(ggplot2::theme_bw())
```

## Overview
This vignette will show a way to analyse longitudinal EQ-5D data using statistical models and functionality from the qalytools package. The goal is to calculate the utility values, identify explanatory variables and calculate the QALY loss. It builds upon the introductory vignette (`vignette("example_analysis")`) and again utilises the synthetic EQ-5D-5L data contained within the package.

```{r setup, message=FALSE}
library(qalytools) # for utility and qaly calculations
library(dplyr)     # for data wrangling
library(tidyr)     # for data wrangling
library(purrr)     # for mapping
library(stringr)   # for string manipulation
library(lme4)      # for modelling
library(ggplot2)   # for plotting
data("EQ5D5L_surveys")
```

## A simple model
The survey includes the age of the participants. For the analysis we group the participants by age ($[20,40), [40,60), [60,+)$). This is common practice in this field, because health outcomes are generally highly dependent on the age of the participants.

```{r thedata}
dat <- 
    qalytools::EQ5D5L_surveys |>
    dplyr::mutate(AgeGroup = cut(age, c(20, 40, 60, Inf), right = FALSE))
```

Next we calculate the utility values using the value the NICE Decision Support Unit (DSU) value set. This maps between
5L and 3L whilst also accounting for the sex and age of respondents.

```{r loaddata}
input_dat <- 
    dat |> 
    qalytools::as_eq5d5l(
        mobility = "mobility",
        self_care = "self_care",
        usual = "usual",
        pain = "pain",
        anxiety = "anxiety",
        respondentID = "respondentID",
        surveyID = "surveyID",
        vas = "vas"
    ) |>
    qalytools::add_utility(type = "DSU", country = "UK", age = "age", sex = "sex")
```

We will be using a mixed effect model to fit the utility values. The model is defined as `value ~ (1 + acute | respondentID) + surveyID + sex + AgeGroup + sex:AgeGroup`. This means that we assume each respondent to have a random effect as well as a random interaction between the first survey after symptoms (acute). Other explanatory variables are the survey, sex and age group (and the interaction between sex and age group). This model was chosen based on our knowledge of the data. For your own dataset it is recommended to choose your own model. For example, by exploring multiple models and choosing the best model (using standard model comparison methods).

To fit our mixed effect model we use the `lme4` package. While the utility value is truncated at 1, it has been shown that assuming a normal distribution and performing a non parametric bootstrap is a valid simplification [@pull_analysis_2010]. We will explore a more complex approach later in this vignette.

```{r lmerModel, fig.cap = "Comparing the modelled utility values with the utility values in the model. Utility values were calculated from a EQ5D5L survey using the DSU method."}
# Label the acute period of the disease
dat <- dplyr::mutate(input_dat, acute = surveyID == "survey02")

# Define the model
model <- .value ~ (1 + acute | respondentID) +
    surveyID + sex + AgeGroup + sex:AgeGroup
fit2 <- lme4::lmer(model, data = dat)

# Compare predictions to the actual value
plot_dat <- 
    dat |>
    dplyr::mutate(Model = predict(fit2)) |>
    dplyr::rename(Actual = .value) |>
    tidyr::pivot_longer(c(Actual, Model), names_to = "type", values_to = "value")

ggplot2::ggplot(plot_dat) +
    ggplot2::geom_line(
        ggplot2::aes(x = surveyID, y = value, group = respondentID, colour = sex),
        alpha = 0.1
    ) +
    ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 45, vjust = 1, hjust = 1)
    ) +
    ggplot2::facet_grid(. ~ type)
```

We will use a non parametric bootstrap to calculate the uncertainty in our model. For this bootstrap we resample (with replacement) from our respondents and fit the model to the resampled data.

```{r bootstrap}
set.seed(1)
# Using 100 samples for illustrative purposes.
# For full analysis you should use >1000
nboot <- 100

respondents_dat <- dplyr::distinct(dat, respondentID)

models <- 
    seq(1, nboot) |> 
    purrr::map(
        purrr:::quietly(function(np_id) {
            # Sample the respondents to include in this bootstrap and give them
            # a unique id (boot_respondent_id)
            boot_dat <- 
                respondents_dat |> 
                dplyr::sample_n(nboot, replace = TRUE) |> 
                dplyr::mutate(boot_respondent_id = dplyr::row_number()) |> 
                dplyr::left_join(dat, by = "respondentID")
          
            fit <- lme4::lmer(
                .value ~ (1 + acute | boot_respondent_id) + surveyID + 
                    sex + AgeGroup + sex:AgeGroup,
                data = boot_dat
            )
          
            list(model = fit, np = as.numeric(np_id), data = boot_dat)
        })
    ) |> 
    purrr::keep(function(lst0) length(lst0$warnings) == 0)
```


Using the bootstrapped models we can capture the uncertainty in the coefficients of our explanatory variables (see figure below). Unsurprisingly, this shows that survey 2 is associated with the lowest utility value. Survey 3, 4 and 5 also had significantly worse outcomes than survey 1, but respondents returned to base line levels of utility from survey 6 onwards. We did not find a significant correlation between age and sex with their utility values, except for 60 and over year old males, who experienced lower utility than the other groups.

```{r coefficients, fig.cap = "Coefficients of the statistical model. Uncertainty was captured by fitting the model to bootstrapped data. The blue colour highlights coefficients that were significant."}
# Gather coefficients from the bootstrapped models
coeff_dat <- 
    models |> 
    purrr::imap(function(model, np_id) {
        summary(model$result$model)$coefficients |> 
            dplyr::as_tibble(rownames = "id") |>
            dplyr::mutate(np = np_id)
    }) |> 
    dplyr::bind_rows()

plot_dat <- 
    coeff_dat |>
    dplyr::group_by(id) |>
    dplyr::summarise(
        quant = c(0.025, 0.25, 0.5, 0.75, 0.975),
        value = quantile(Estimate, quant),
        Significant = all(value < 0) || all(value > 0),
        .groups = "drop"
    ) |> 
    tidyr::pivot_wider(names_from = quant, values_from = value) |> 
    # Cleanup names of the coefficients
    dplyr::mutate(idname = stringr::str_replace(id, "(sex|surveyID)", "")) |>
    dplyr::filter(idname != "(Intercept)")

ggplot2::ggplot(data = plot_dat) +
    ggplot2::geom_linerange(
        ggplot2::aes(x = idname, ymin = `0.025`, ymax = `0.975`, colour = Significant),
        size = 1
    ) +
    ggplot2::geom_linerange(
        ggplot2::aes(x = idname, ymin = `0.25`, ymax = `0.75`, colour = Significant),
        size = 2
    ) +
    ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 45, vjust = 1, hjust = 1),
        axis.title.x = ggplot2::element_blank(),
        legend.position = "none"
    )
```

Finally, we use the model to generate utility values and calculate the QALY loss by age group. QALY loss compared to full health does increase with age, but when compared to the baseline health (survey 1) it stayed relatively stable.

```{r qaly}
# Calculate the QALY values for the different bootstrapped models
qaly_dat <- 
    models |>
    purrr::imap(function(lst0, np_id) {
        lst0$result$data |>
            dplyr::mutate(pvalue = predict(lst0$result$model)) |>
            # Convert to utility
            qalytools::new_utility(
                respondentID = "boot_respondent_id",
                surveyID = "surveyID",
                country = ".utility_country",
                type = ".utility_type",
                value = "pvalue"
            ) |>
            qalytools::validate_utility() |>
            # Calculate the qaly for based on the utility values
            qalytools::calculate_qalys(
                baseline_survey = "survey01",
                time_index = "time_index"
            ) |>
            dplyr::mutate(np = as.numeric(np_id)) |>
            dplyr::left_join(
                dplyr::distinct(lst0$result$data, boot_respondent_id, AgeGroup, sex),
                by = "boot_respondent_id"
            )
    }) |> 
    dplyr::bind_rows()
```

```{r qalyFig, fig.cap = "QALY values based on the model. The cross is the mean QALY based on the data. The uncertainty represents the uncertainty in the mean QALY loss according to the model"}
plot_dat <- 
    qaly_dat |>
    dplyr::group_by(.qaly, AgeGroup, np) |>
    dplyr::summarise(value = mean(.value)) |>
    dplyr::group_by(.qaly, AgeGroup) |>
    dplyr::summarise(
        quant = c(0.025, 0.25, 0.5, 0.75, 0.975),
        value = quantile(value, quant),
        groups = "drop"
    ) |>
    tidyr::pivot_wider(names_from = quant, values_from = value) |> 
    dplyr::filter(.qaly != "raw")

# Include qalys calculated from the raw data
dat0 <- 
    input_dat |>
    qalytools::calculate_qalys(
        baseline_survey = "survey01",
        time_index = "time_index"
    ) |>
    dplyr::filter(.qaly != "raw") |>
    dplyr::left_join(distinct(dat,respondentID, AgeGroup, sex)) |>
    dplyr::group_by(AgeGroup, .qaly) |>
    dplyr::summarise(mean = mean(.value), .groups = "drop")

ggplot2::ggplot(data = plot_dat) +
    ggplot2::geom_linerange(
        ggplot2::aes(x = AgeGroup, ymin = `0.025`, ymax = `0.975`, colour = .qaly, group = .qaly),
        position = ggplot2::position_dodge2(0.5)
    ) +
    ggplot2::geom_linerange(
        ggplot2::aes(x = AgeGroup, ymin = `0.25`, ymax = `0.75`, colour = .qaly, group = .qaly),
        position = position_dodge2(0.5),
        size = 2
    ) +
    ggplot2::geom_point(
        data = dat0, 
        ggplot2::aes(x = AgeGroup, y = mean, group = .qaly),
        position = ggplot2::position_dodge2(0.5),
        size = 3,
        shape = 4
    ) +
    ggplot2::expand_limits(y = 0)
```

## A better model

```{r brmsanalysis}
# Use multiple cores for brms
options(mc.cores = 4)
library(brms)

input_dat |>
  dplyr::mutate(acute = surveyID == "survey02") -> dat

brm(bf(.value ~ (1 + acute | respondentID) + surveyID + sex + AgeGroup +
       sex:AgeGroup), data = dat, family = gaussian(), iter = 2000,
    # Suppress messages for the vignette
    silent = 2, refresh = 0, open_progress = F) -> brm_model

# Draw posterior samples to use in qaly calculation. We ignore the residual errors/measurement
# when calculating the QALY
brms::posterior_epred(brm_model, newdata = dat, ndraws = 1000) -> posterior_dat

# For each posterior sample, calculate the qaly loss
purrr::map(seq_len(nrow(posterior_dat)), function(np_id) {
  dat |> dplyr::mutate(pvalue = posterior_dat[as.numeric(np_id), ]) |>
    qalytools::new_utility(respondentID = "respondentID",
                             surveyID = "surveyID",
                             country = ".utility_country",
                             type = ".utility_type",
                             value = "pvalue") |>
    qalytools::validate_utility() |>
    qalytools::calculate_qalys(baseline_survey = "survey01",
                               time_index = "time_index") |>
    dplyr::mutate(np = as.numeric(np_id))
}) |> dplyr::bind_rows() |>
    dplyr::left_join(dat |>
      dplyr::select(respondentID, AgeGroup, sex) |>
      dplyr::distinct()) -> qaly_dat
```

```{r brmsModel, fig.cap = "Comparing the modelled utility values, using the bayesian model, with the utility values in the model. Utility values were calculated from a EQ5D5L survey using the DSU method."}
# For comparison, plot posterior using one posterior sample
dat |> dplyr::mutate(pvalue = posterior_dat[1, ]) |>
  qalytools::new_utility(respondentID = "respondentID",
                           surveyID = "surveyID",
                           country = ".utility_country",
                           type = ".utility_type",
                           value = "pvalue") |>
  qalytools::validate_utility() -> dat0
dat0 |>
  dplyr::rename(Actual = .value, Model = pvalue) |>
  tidyr::gather(type, value, Actual, Model) -> plot_dat

ggplot(plot_dat) +
  geom_line(aes(x = surveyID, y = value, group = respondentID,
                colour = sex), alpha = 0.1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  facet_grid(. ~ type)
```

```{r coefficientsBrms, fig.cap = "Coefficients of bayesian model. Uncertainty was captured by fitting the model to bootstrapped data. The blue colour highlights coefficients that were significant. Note that because we transformed our utility variable, the sign of these are the opposite of the previous results."}
fixef(brm_model, probs = c(0.025, 0.25, 0.5, 0.75, 0.975)) -> dat0
dat0 |>
  as_tibble() |>
  dplyr::mutate(id = rownames(dat0)) |>
  dplyr::mutate(Significant = (`Q2.5` < 0 & `Q97.5` < 0) |
                (`Q2.5` > 0 & `Q97.5` > 0)) |>
  # Cleanup names of the coefficients
  dplyr::mutate(idname = stringr::str_replace(id, "(sex|surveyID)", "")) |>
  dplyr::filter(idname != "Intercept") -> plot_dat

ggplot(data = plot_dat) +
  geom_linerange(aes(x = idname, ymin = `Q2.5`, ymax = `Q97.5`,
                     colour = Significant), size = 1) +
  geom_linerange(aes(x = idname, ymin = `Q25`, ymax = `Q75`,
                     colour = Significant), size = 2) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.title.x = element_blank(),
        legend.position = "none")
```

```{r qalyFigBrms, fig.cap = "QALY values based on the bayesian model. The cross is the mean QALY based on the data. The uncertainty represents the uncertainty in the mean QALY loss according to the model"}
qaly_dat |>
  dplyr::group_by(.qaly, AgeGroup, np) |>
  dplyr::summarise(value = mean(.value)) |>
  dplyr::group_by(.qaly, AgeGroup) |>
  dplyr::summarise(quant = c(0.025, 0.25, 0.5, 0.75, 0.975),
                   value = quantile(value, quant)) |>
  dplyr::ungroup() |>
  tidyr::spread(quant, value) |>
  dplyr::filter(.qaly != "raw") -> plot_dat

# Include qalys calculated from the raw data
input_dat |>
  qalytools::calculate_qalys(baseline_survey = "survey01", time_index = "time_index") |>
  dplyr::filter(.qaly != "raw") |>
  dplyr::left_join(dat |>
    dplyr::select(respondentID, AgeGroup, sex) |>
    dplyr::distinct()) |>
  dplyr::group_by(AgeGroup, .qaly) |>
  dplyr::summarise(mean = mean(.value)) |>
  dplyr::ungroup() -> dat0
ggplot(data = plot_dat) +
  geom_linerange(aes(x = AgeGroup, ymin = `0.025`, ymax = `0.975`,
                     colour = .qaly, group = .qaly),
                 position = position_dodge2(0.5)) +
  geom_linerange(aes(x = AgeGroup, ymin = `0.25`, ymax = `0.75`,
                     colour = .qaly, group = .qaly),
                 position = position_dodge2(0.5),
                 size = 2) +
  geom_point(data = dat0, aes(x = AgeGroup, y = mean, group = .qaly),
                 position = position_dodge2(0.5), size = 3,
                 shape = 4) +
  expand_limits(y = 0)
```

# References

