---
title: "surveyTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{surveyTools}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    fig.width = 8,
    fig.height = 5
)
```

***surveyTools*** provides a simple and intuitive user interface for the
analysis of [EQ-5D](https://euroqol.org/eq-5d-instruments) surveys. It builds
upon the [eq5d package](https://cran.r-project.org/package=eq5d) to facilitate
the calculation of QALY values and other related metrics across multiple
surveys.

The package provides a range of functions:

- Constructors (and validators) for EQ5D data frame subclasses (EQ5D3L, EQ5D5L
and EQ5D): `new_eq5d3l()` (`validate_eq5d3l()`), `new_eq5d5l()`
(`validate_eq5d5l()`) and `new_eq5dy()` (`validate_eq5dy()`).
- User-friendly coercion functions `as_eq5d3l()`, `as_eq5d5l()` and 
`as_eq5dy()`.
- The calculation of utility values based on a range of different value sets.
This functionality is provided via the `calculate_utility()`, `add_utility()`
and `available_valuesets()` functions which are wrappers around the
[eq5d](https://cran.r-project.org/package=eq5d) package.
- The calculation of different Quality of Life Years (QALY) metrics including
unadjusted 'raw' values, and the disutility from both perfect health and,
optionally, a specified baseline. See `calculate_qalys()`.
- The calculation of the Paretian Classification of Health Change (PCHC) in an
individual's health state between two surveys via `calculate_pchc()` (again
wrapping the [eq5d](https://cran.r-project.org/package=eq5d) package).
- Easy calculation of responses with a health limitation (i.e. a non-one
response in one of the dimensions) via `calculate_limitation()`.


## The EQ5D object class

Table: Example data

| surveyID| respondentID| time_index| mobility| self_care| usual| pain| anxiety| vas|sex    | age|
|--------:|------------:|----------:|--------:|---------:|-----:|----:|-------:|---:|:------|---:|
|        .|            .|         . |        .|         .|     .|    .|       .|   .|.      |   .|
|        1|            1|         30|        3|         2|     2|    1|       1|  87|Female |  25|
|        1|            2|         30|        2|         2|     2|    2|       2|  65|Male   |  30|
|        1|            3|         30|        1|         3|     3|    1|       3|  76|Male   |  39|
|        1|            4|         30|        1|         2|     3|    2|       4|  55|Female |  60|
|        1|            5|         30|        2|         1|     4|    1|       1|  80|Male   |  28|
|        1|            6|         30|        1|         3|     2|    1|       1|  83|Male   |  59|
|        .|            .|         . |        .|         .|     .|    .|       .|   .|.      |   .|

We define an `EQ5D` object as a table containing data in long format that meets
the following, additional, criteria:

-   It contains columns that represent dimensions from the EQ5D survey
specification as well as a column representing the Visual Analogue Score
(columns 'mobility', 'self_care', 'usual', 'pain', 'anxiety' and 'vas' in the
example data above)

-   Dimension values must be whole numbers, bounded between 1 and 3 for EQ5D3L
and EQ5D3Y surveys or bounded between 1 and 5 for EQ5D5L surveys

-   It contains a column that acts as a unique respondent identifier
('respondentID') and another that identifies different surveys over time
('surveyID'). Together these should uniquely identify a response and no
combination of these should be duplicated within the data frame.

-   It contains a column that provides the relative time (e.g. days/weeks) of a
response within the survey framework ('time_index').

In *{surveyTools}* we implement these objects as a data frame subclass and
provide functionality for users to coerce input to this representation via the
`as_eq5d3l()`, `as_eq5d5l()` and `as_eq5dy()` functions.


## Usage

```{r}
library(surveyTools)
library(ggplot2)

# Example EQ5D5L data 
data("EQ5D5L_surveys")
str(EQ5D5L_surveys)
```

Before coercing our data to an EQ5D5L object we must ensure that our surveys
have an explicit ordering. We do this by converting the `surveyID` variable to
an ordered factor prior to call `as_eq5d5l()`.

```{r}
dat <- EQ5D5L_surveys
dat$surveyID <- as.factor(dat$surveyID)

dat <- as_eq5d5l(
    dat,
    mobility = "mobility",
    self_care = "self_care",
    usual = "usual",
    pain = "pain",
    anxiety = "anxiety",
    respondentID = "respondentID",
    surveyID = "surveyID",
    vas = "vas",
    time_index = "time_index"
)

dat
```

To obtain a quick overview of the data we can call `summary()`. By default this
returns the output as a list of data frames, showing frequency counts and
proportions, split by `surveyID`

```{r}
head(summary(dat), n = 2)
```

Alternatively we can set the parameter `tidy` to TRUE and obtain the summary
data in a "tidy" (long) format.

```{r}
summary(dat, tidy = TRUE)
```

If we try and coerce an object that does not meet the required criteria, the
function will error with an informative message.

```{r, error=TRUE}
# Incorrect bounds for dimension values will cause an error
as_eq5d3l(
    EQ5D5L_surveys,
    mobility = "mobility",
    self_care = "self_care",
    usual = "usual",
    pain = "pain",
    anxiety = "anxiety",
    respondentID = "respondentID",
    surveyID = "surveyID",
    vas = "vas",
    time_index = "time_index"
)

# Dimension values that are not whole numbers will cause an error
EQ5D5L_surveys[1,5] <- 1.5
as_eq5d5l(
    EQ5D5L_surveys,
    mobility = "mobility",
    self_care = "self_care",
    usual = "usual",
    pain = "pain",
    anxiety = "anxiety",
    respondentID = "respondentID",
    surveyID = "surveyID",
    vas = "vas",
    time_index = "time_index"
)
```

### Available value sets

Once we have created our EQ5D objects we can think about which value sets we
would to use for the calculation of utility values and subsequent QALY metrics.
We can obtain a list of compatible value sets (across countries and type) by
passing our EQ5D object directly to the `available_valuesets()` function or by
passing a comparable string.

For EQ5D3L inputs the type can be:

-   "TTO", the time trade-off valuation technique;

-   "VAS", the visual analogue scale valuation technique;

-   "RCW", a reverse crosswalk conversion to EQ5D5L values; or

-   "DSU", the NICE Decision Support Unit's model that allows mappings on to
EQ5D5L values accounting for both age and sex.

For EQ5D5L inputs this can be:

-   "VT", value sets generated via a EuroQol standardised valuation study
protocol;

-   "CW", a crosswalk conversion EQ5D3L values; or

-   "DSU", the NICE Decision Support Unit's model that allows mappings on to
EQ5D5L values accounting for both age and sex.

Note that this function is purely a convenience wrapper around the
`eq5d::valuesets()` function. It will return a data frame with columns
representing the EQ5D version, the value set country and the value set type.

```{r}
vs <- available_valuesets(dat)
head(vs, 10)
identical(vs, available_valuesets("eq5d5l"))
identical(vs, available_valuesets("EQ5D5L"))
identical(vs, available_valuesets("EQ-5D-5L"))
```

### Calculating utility values

To calculate the utility values we call `calculate_utility` on our EQ5D object
with additional arguments specifying the countries and type we are interested
in. Note that the function will error if the type and country arguments are of
different length, with an exception made for inputs of length one that will be
recycled to the necessary length.

```{r}
# Example 1
vs_dat <- vs[vs$Country %in% c("England", "Ireland", "Germany", "UK", "USA"),]
vs_dat <- vs_dat[vs_dat$Type %in% c("VT", "CW"),]
calculate_utility(dat, type = vs_dat$Type, country = vs_dat$Country)

# Example 2 - cross walk comparison
vs_dat <- vs_dat[vs_dat$Country %in% c("England", "UK"),]
util <- calculate_utility(dat, type = vs_dat$Type, country = vs_dat$Country)

# plot the results
util$fill <- paste(util$.utility_country, util$.utility_type)
ggplot(util, aes(x = surveyID, y = .value, fill = fill)) +
    geom_boxplot(lwd = 1, outlier.shape = 4) +
    stat_summary(
        mapping = aes(group = .utility_country),
        fun = mean,
        geom = "point",
        position = position_dodge(width = 0.75),
        shape = 21,
        color = "black",
        fill = "white"
    ) +
    theme_light() +
    geom_hline(yintercept = 0, linetype = "longdash", size = 0.6, color = "grey30") +
    scale_y_continuous(n.breaks = 10, expand = c(0.005, 0.005)) +
    scale_x_discrete(name = "value set")
```

### QALY calculations

Quality of life years can be calculated directly from utility values. By
default, two different metrics are provided. Firstly, a "raw" value which is
simply the area under the utility curve appropriately scaled and, secondly, a
value which represents the loss from full health.

```{r}
qalys <- 
    dat |> 
    calculate_utility(type = "VT", country = c("Denmark", "France")) |> 
    calculate_qalys()

subset(qalys, .qaly=="raw")
subset(qalys, .qaly=="loss_vs_fullhealth")
```

### Comparison to a baseline

`calculate_qalys()` also allows us to calculate the loss from a specified
baseline in one of two ways. Firstly, a character string `baseline_survey`
argument can be placed which matches a survey present in the utility data. If
the argument is passed as such then the utility values from the specified survey
are used to calculate the loss. Note that the survey is still included in the
raw, unadjusted calculation, prior to the calculation of loss.

```{r}
# Reload the example data and combine with some baseline measurements
data("EQ5D5L_surveys")
dat <- 
    EQ5D5L_surveys |> 
    transform(surveyID = as.factor(surveyID)) |> 
    as_eq5d5l(
        mobility = "mobility",
        self_care = "self_care",
        usual = "usual",
        pain = "pain",
        anxiety = "anxiety",
        respondentID = "respondentID",
        surveyID = "surveyID",
        vas = "vas",
        time_index = "time_index"
    )

calculate_utility(dat, type = "VT", country = c("Denmark", "France")) |> 
    calculate_qalys(baseline_survey = "survey01") |> 
    subset(.qaly == "loss_vs_baseline")
```

Alternatively the `baseline_survey` argument can be specified as a data frame
with a column corresponding to the respondentID and another representing the
associated utility. Optionally columns corresponding to the utility country and
utility type can be included to allow more granular comparisons. Note that for
this specification of baseline, it is **not** included in the unadjusted, raw,
calculation.

```{r}
split_dat <- split(dat, dat$surveyID=="survey01")
surveys <- split_dat[[1]]
baseline <- split_dat[[2]]
utility_dat <- calculate_utility(surveys,
                                 type = "VT",
                                 country = c("Denmark", "France"))
baseline_utility <- 
    baseline |> 
    calculate_utility(type = "VT", country = c("Denmark", "France")) |> 
    subset(select=c(respondentID,.utility_country,.utility_type,.value))

calculate_qalys(utility_dat, baseline_survey = baseline_utility) |> 
    subset(.qaly == "loss_vs_baseline")
```

### Calculate limitations

```{r}
(limitation <- calculate_limitation(dat))

ggplot(limitation, aes(x = surveyID, y = without_limitation, group = dimension)) +
    geom_line(aes(colour = dimension)) +
    theme_light() +
    scale_y_continuous(n.breaks = 10, expand = c(0.005, 0.005), limits = c(0, 1)) +
    scale_x_discrete() +
    scale_fill_discrete(name = "dimension") +
    ylab("Without limitation")
```

### Paretian Classification of Health Change (PCHC)

```{r}
data("eq5d3l_example")
dat <- as_eq5d3l(
    eq5d3l_example,
    respondentID = "respondentID",
    surveyID = "surveyID",
    time_index = "time",
    mobility = "MO",
    self_care = "SC",
    usual = "UA",
    pain = "PD",
    anxiety = "AD",
    vas = "vas",
    drop = FALSE # ensure we do not drop additional columns
) 
grp1 <- subset(dat, Group == "Group1")
grp2 <- subset(dat, Group == "Group2")
calculate_pchc(grp1, grp2)
calculate_pchc(grp1, grp2, by.dimension = TRUE)
```
